Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes

### 2020-09-30 GWLDCI.tex
line about first paper not prerequisite in paper 1
stab at report M

\title{Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes}
\author{Susama Agarwala\thanks{SA was partially supported by an Office of Naval Research grant.}, Si\^an Fryer, and Karen Yeats\thanks{KY is supported by an NSERC Discovery grant, by the Canada Research Chair program, and though some the time this work was developed also by a Humboldt Fellowship from the Alexander von Humboldt foundation.}}
%\date{}

@@

The second paper \cite{generalcombinatoricsII} in this series identifies precisely which positroid cell is associated to a given Wilson loop diagram, by giving an algorithm for passing from a Wilson loop diagram to the Grassmann necklace of the associated positroid cell. In this manner we give a direct path from Wilson loop diagram to positroid cell, circumventing the painstaking process of examining sets of bases. With this identification in hand, we also prove that any Wilson loop diagram with $k$ propagators corresponds to a $3k$ dimensional positroid cell, and characterize the volume forms that Wilson loop diagrams associate to each positroid cell.

@@

\begin{dfn} \label{VPropdfn}
Let $W = (\cP, [n])$.
\begin{enumerate}
\item For $p = (i,j) \in \cP$, define $V(p) = \{i, i+1, j, j+1\}$ to be the set of vertices supporting $p$. Similarly, for $P \subseteq \cP$ we define the set $V(P) = \cup_{p \in P} V(p)$ to be the vertex support of $P$.
\item For $U \subseteq [n]$, write $\Prop(U) = \{ p \in \cP  \ | \ V(p) \cap U \neq \emptyset \} $ to denote the set of propagators which are at least partially supported on $U$.
\item For $P \subseteq \cP$, define the {\em propagator flat}\footnote{This choice of terminology is justified in Lemma~\ref{lem facts about WLD matroids} below.} $F(P) = V(P^c)^c$ to be the set of vertices in $[n]$ that do not support any propagators outside of the set $P$. Here, as throughout, we are using the superscript $c$ for the set theoretic complement.
\item The set of vertices that do not support any propagators is denoted $F(\emptyset)$. Vertices in this set are called {\em non-supporting}.  
\end{enumerate}
\end{dfn}

@@

\begin{dfn}\label{admisdfn}
A Wilson loop diagram $W = (\cP,V)$ is {\em admissible} if it satisfies the following three conditions: \begin{enumerate}
\item $|V| \geq |\cP| + 4.$
\item There does not exist a (non-empty) set of propagators $Q \subseteq \cP$ such that $|V(Q)| < |Q| + 3$.
\item There does not exist a pair of propagators $(i_p,j_p),(i_q,j_q) \in \cP$ such that $i_p < i_q < j_p <j_q$.
\end{enumerate}
A Wilson loop diagram is called {\em weakly admissible} if the second and third conditions hold.
 \end{dfn}

@@

There are a few things to note here.  First, what we call weakly admissible here is called admissible in \cite{wilsonloop}.  Many of the results below still apply when the hypothesis of admissibility is changed to weak admissibility.

@@

Finally, we describe several important types of matroids. A matroid of rank $k$ with a ground set of size $n$ is called {\em realizable} if there exists some $A$ in the $k\times n$ real Grassmannian $\mathbb{G}_{\mathbb{R}}(k,n)$ whose non-zero $k\times k$ minors are exactly those with columns indexed by an element of $\cB$. If all the maximal minors of the matrix are nonnegative then the matroid realized by this matrix is called a {\em positroid}.  In other words a positroid is a matroid\footnote{Note that matroid isomorphism allows arbitrary permutations of the ground set, while positroid isomorphism only allows cyclic permutations in order to preserve the nonnegativity of the minors.  In particular, the property of being a positroid is not preserved under matroid isomorphism.} that is realized by an element of the totally nonnegative Grassmannian $\Gr(k,n)$. Finally, a {\em uniform matroid} of rank $r$ is a matroid in which any set of size $\leq r$ is independent.

@@

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} Any admissible Wilson loop diagram $W =(\cP, [n])$ defines a matroid $M(W)$ with ground set $[n]$. The independent sets are exactly those subsets $V \subseteq [n]$ such that $\nexists U\subseteq V$ satisfying $|\Prop(U)| < |U|$. \label{thm:WLDmatroid}\end{thm}
In other words, the independent sets of $M(W)$ correspond to the sets of vertices in $W$ with ``enough propagators sufficiently well distributed'' on them: not only should $V$ itself support at least as many propagators as the number of vertices it contains, but the same should be true of any subset of $V$ as well.

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note from Definition~\ref{def:CWmatrix} that the vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$. Therefore, for any subset of vertices $V \subseteq [n]$ we can consider the restriction $C(W)|_V$ of $C(W)$, which consists of the columns corresponding to $V$. Since the non-zero entries of $C(W)$ are independent real variables, if $C(W)|_V$ has fewer than $|V|$ rows of non-zero entries then it does not have full rank. This is equivalent to saying that if $V$ is a set of columns of $C(W)$ with $|\Prop(V)| < |V|$, then $V$ is a dependent set. Any set of columns containing such a $V$ cannot be independent. Comparing this to Theorem \ref{thm:WLDmatroid}, we see that $M(W)$ is also the matroid realized by $C(W)$.

@@

\section{Matroids, triangulations, and equivalence classes of Wilson loop diagrams}\label{sec equiv}

@@

In this section we develop this analysis further, showing the converse of this result from loc. cit. (see Theorem \ref{same matroid iff equiv}). We also enumerate the number of diagrams that map to a particular matroid in Corollary \ref{number of equiv diagrams}. To do this, we establish a connection between Wilson loop diagrams and polygon dissections in subsection \ref{sec: polygon partitions}, and uncover some matroidal properties of exact subdiagrams in subsection \ref{sec: exact diagram matroidal props}. This includes the fact that exact subdiagrams correspond to uniform matroids (Theorem \ref{exactuniformthm}), and that the associated matroid of an exact subdiagram can be written as as a contraction by the complementary propagator flat (Theorem \ref{exact diagrams contractions}).

@@

Since Corollary \ref{uniqueproppartitioncor} allows us to decompose any admissible Wilson loop diagram into a collection of maximal exact subdiagrams, we are motivated to examine the matroid properties of exact subdiagrams more closely. In this section, we prove two main results:
\begin{enumerate}
\item The matroid assoicated to an exact subdiagram of $W$ can be written as a contraction of the matroid $M(W)$ by the complementary propagator flat (Theorem \ref{exact diagrams contractions}). 
\item The matroid associated to an exact subdiagram is uniform (Theorem \ref{exactuniformthm}).
\end{enumerate}

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram of $W$, then $F(P^c)$ is a dependent flat in $W$.
\item If $(P,V(P))$ is a maximal exact subdiagram of $W$, then $F(P^c)$ is a cyclic flat in $W$.
\item If $(P,V(P))$ is an exact subdiagram in $W$, then $\rk(F(P^c)) = |P^c|$.
\end{enumerate} 
\end{lem}

@@

\begin{thm} \label{exact diagrams contractions}
If $(P, V(P))$ is an exact subdiagram of of $W$, the matroid of $(P,V(P))$ is the contraction of $M(W)$ by the propagator flat $F(P^c)$, i.e. \bas M\big((P, V(P))\big) = M(W)  / F(P^c) \;.\eas 
\end{thm}

@@

\begin{thm}\label{same matroid iff equiv}
Let $W= (\cP, [n])$ and $W'= (\cP', [n])$ be two Wilson loop diagrams. They define the same matroid (positroid) if and only if $W \sim W'$.
\end{thm}

@@

\begin{dfn}\label{def:associahedron}
Fix $n$ and consider dissections of a convex $n$-gon.  The $n-1$ {\em associahedron} can be defined in terms of the dissections of this $n$-gon: the vertices of the associahedron are the triangulations of the $n$-gon, while the $k$-dimensional faces are the dissections which are $k$ diagonals away from being a triangulation.  One face is included in another if the first dissection is a refinement of the second. 
\end{dfn}

@@

\begin{dfn}\label{def:secondary polytope}
  Let $x_1, \ldots, x_n$ be the corners of a convex $n$-gon in $\mathbb{R}^2$, and let $T$ be the set of triangulations of this $n$-gon.  For each $t\in T$, define $s_t$ to be the point in $\mathbb{R}^n$ whose $i$th coordinate is the sum of the areas of all triangles of $t$ which are incident to $x_i$.  Let $A_n$ be the convex hull of the $s_t$; then $A_n$ is a realization of the $n-1$ associahedron (\cite[Example 9.11]{Ziegler}).
\end{dfn}
Note that this construction works for any convex $n$-gon, but these realizations are all equally good for our purposes, and so we fix $x_1, \ldots, x_n$ defining a particular convex $n$-gon from now on.

@@

\begin{proof}
  Since $D_1$ and $D_2$ do not have the same set of non-triangle faces, there must be some diagonal $d$ of $D_1$ that crosses through a non-triangle face $f$ of $D_2$.  If that face of $D_2$ has degree at least 5 then it must contain four vertices in the configuration described in the statement.

>>>>

\title{Combinatorics of the geometry of Wilson loop diagrams I: equivalence classes via matroids and polytopes}
\author{Susama Agarwala\thanks{SA was partially supported by an Office of Naval Research grant.}, Si\^an Fryer, and Karen Yeats\thanks{KY is supported by an NSERC Discovery grant, by the Canada Research Chair program, though some of this work was developed also by a Humboldt Fellowship from the Alexander von Humboldt foundation.}}
%\date{}

@@

The second paper \cite{generalcombinatoricsII} in this series identifies precisely which positroid cell is associated to a given Wilson loop diagram, by giving an algorithm for passing from a Wilson loop diagram to the Grassmann necklace of the associated positroid cell. In this manner we give a direct path from Wilson loop diagram to positroid cell, circumventing the painstaking process of examining sets of bases. With this identification in hand, we also prove that any Wilson loop diagram with $k$ propagators corresponds to a $3k$ dimensional positroid cell, and characterize the volume forms that Wilson loop diagrams associate to each positroid cell. The results of the first paper are not prerequisite for the second paper in this series.

@@

\begin{dfn} \label{VPropdfn}
Let $W = (\cP, [n])$. In this definition and in the sequel, for any $P \subseteq \cP$ or $U \subseteq [n]$, we denote $P^c = \cP \setminus P$ and $U^c = [n] \setminus U$.
\begin{enumerate}
\item For $p = (i,j) \in \cP$, define $V(p) = \{i, i+1, j, j+1\}$ to be the set of vertices supporting $p$. Similarly, for $P \subseteq \cP$ we define the set $V(P) = \cup_{p \in P} V(p)$ to be the vertex support of $P$.
\item For $U \subseteq [n]$, write $\Prop(U) = \{ p \in \cP  \ | \ V(p) \cap U \neq \emptyset \} $ to denote the set of propagators which are at least partially supported on $U$.
\item For $P \subseteq \cP$, define the {\em propagator flat}\footnote{This choice of terminology is justified in Lemma~\ref{lem facts about WLD matroids} below.} $F(P) = V(P^c)^c$ to be the set of vertices in $[n]$ that do not support any propagators outside of the set $P$. Here, as throughout, we are using the superscript $c$ for the set theoretic complement.
\item The set of vertices that do not support any propagators is denoted $F(\emptyset)$. Vertices in this set are called {\em non-supporting}.  
\end{enumerate}
\end{dfn}

@@

\begin{dfn}\label{admisdfn}
A Wilson loop diagram $W = (\cP,V)$ is {\em admissible} if it satisfies the following three conditions: \begin{enumerate}
\item $|V| \geq |\cP| + 4.$
\item For any non-empty set of propagators, $Q \subseteq \cP$, $Q \neq \emptyset$, one has $|V(Q)| \geq |Q| + 3$.
\item There does not exist a pair of propagators $(i_p,j_p),(i_q,j_q) \in \cP$ such that $i_p < i_q < j_p <j_q$.
\end{enumerate}
A Wilson loop diagram is called {\em weakly admissible} if the second and third conditions hold.
 \end{dfn}

@@

There are a few things to note here.  First, what we call weakly admissible here is called admissible in \cite{wilsonloop}.  \hlfix{Many of the results below still apply when the hypothesis of admissibility is changed to weak admissibility.}{Propagate "weakly" through document}

@@

Finally, we describe several important types of matroids. A matroid of rank $k$ with a ground set of size $n$ is called {\em realizable} if there exists some matrix $A$ in the $k\times n$ real Grassmannian $\mathbb{G}_{\mathbb{R}}(k,n)$ whose non-zero $k\times k$ minors are exactly those with columns indexed by an element of $\cB$. If all the maximal minors of the matrix are nonnegative then the matroid realized by this matrix is called a {\em positroid}.  In other words a positroid is a matroid\footnote{Note that matroid isomorphism allows arbitrary permutations of the ground set, while positroid isomorphism only allows cyclic permutations in order to preserve the nonnegativity of the minors.  In particular, the property of being a positroid is not preserved under matroid isomorphism.} that is realized by an element of the totally nonnegative Grassmannian $\Gr(k,n)$. Finally, a {\em uniform matroid} of rank $r$ is a matroid in which any set of size $\leq r$ is independent.

@@

\begin{thm} \label{thm WLD defines matroid} \cite[Theorem 3.6]{wilsonloop} Any admissible Wilson loop diagram $W =(\cP, [n])$ defines a matroid $M(W)$ with ground set $[n]$. The independent sets are exactly those subsets $V \subseteq [n]$ such that for every subset of vertices, $U\subseteq V$, supports at least as many propagators as vertices in the set, $|\Prop(U)| \geq |U|$. \label{thm:WLDmatroid}\end{thm}
In other words, the independent sets of $M(W)$ correspond to the sets of vertices in $W$ with ``enough propagators sufficiently well distributed'' on them: not only should $V$ itself support at least as many propagators as the number of vertices it contains, but the same should be true of any subset of $V$ as well.

Throughout, we take the {\em matroid defined by $W$} to be the matroid $M(W)$ of Theorem \ref{thm WLD defines matroid}. Note from Definition~\ref{def:CWmatrix} that the vertices of the diagram $W$ correspond to columns of the associated matrix $C(W)$. Therefore, for any subset of vertices $V \subseteq [n]$ we can consider the restriction $C(W)|_V$ of $C(W)$, which consists of the columns corresponding to $V$. Since the non-zero entries of $C(W)$ are independent real variables, if $C(W)|_V$ has fewer than $|V|$ rows of non-zero entries then it does not have full rank. This is equivalent to saying that if $V$ is a set of columns of $C(W)$ with $|\Prop(V)| < |V|$, then $V$ is a dependent set. Any set of columns containing such a $V$ cannot be independent. Comparing this to Theorem \ref{thm:WLDmatroid}, we see that $M(W)$ is also the matroid realized by $C(W)$. As mentioned above, Theorem [3.38] of \cite{wilsonloops} shows that the matroids $M(W)$ is a positroid. Furthermore, Theorem 5.1 \cite{basisshapeloci} shows that the space parametrized by the matrices $C(W)$ matches the positroid cells defined by $M(W)$, up to a set of measure $0$. Note, this does not imply that every possible specializaton of $C(W)$ to a real valued matrix leads to a postroid, only that the set of values that do define a positroid is volume filling in $\R^{4k}$. \todo{What is [6] referred to by reviewer M?}

@@

\todo{Need to address the summary comment to 3.1 in review M}

\section{Matroids, triangulations, and equivalence classes of Wilson loop diagrams\label{sec equiv}} \todo{rephrase in terms of weakly.}

@@

In this section we develop this analysis further, showing the converse of this result from loc. cit. (see Theorem \ref{same matroid iff equiv}). We also enumerate the number of diagrams that map to a particular matroid in Corollary \ref{number of equiv diagrams}. To do this, we establish a connection between Wilson loop diagrams and polygon dissections in subsection \ref{sec: polygon partitions}, and uncover some matroidal properties of exact subdiagrams in subsection \ref{sec: exact diagram matroidal props}. This includes the fact that exact subdiagrams correspond to uniform matroids (Theorem \ref{exactuniformthm}), and that the associated matroid of an exact subdiagram can be written as a contraction by the complementary propagator flat (Theorem \ref{exact diagrams contractions}).

@@

Since Corollary \ref{uniqueproppartitioncor} allows us to decompose any admissible Wilson loop diagram into a collection of maximal exact subdiagrams, we are motivated to examine the matroid properties of exact subdiagrams more closely. In this section, we prove two main results:
\begin{enumerate}
\item The matroid associated to an exact subdiagram of $W$ can be written as a contraction of the matroid $M(W)$ by the complementary propagator flat (Theorem \ref{exact diagrams contractions}). 
\item The matroid associated to an exact subdiagram is uniform (Theorem \ref{exactuniformthm}).
\end{enumerate}

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram of $W$, then $F(P^c)$ is a dependent flat in $W$.
\item If $(P,V(P))$ is a maximal exact subdiagram of $W$, then $F(P^c)$ is a cyclic flat in $W$.
\item If $(P,V(P))$ is an exact subdiagram of $W$, then $\rk(F(P^c)) = |P^c|$.
\end{enumerate} 
\end{lem}

@@

\begin{thm} \label{exact diagrams contractions}
If $(P, V(P))$ is an exact subdiagram of $W$, the matroid of $(P,V(P))$ is the contraction of $M(W)$ by the propagator flat $F(P^c)$, i.e. \bas M\big((P, V(P))\big) = M(W)  / F(P^c) \;.\eas 
\end{thm}

@@

\begin{thm}\label{same matroid iff equiv}
Let $W= (\cP, [n])$ and $W'= (\cP', [n])$ be two weakly admissible Wilson loop diagrams. They define the same matroid (positroid) if and only if $W \sim W'$.
\end{thm}

@@

\begin{dfn}\label{def:associahedron} \todo{Karen, see note bottom of page 2 in report M}
Fix $n$ and consider dissections of a convex $n$-gon.  The $n-1$ {\em associahedron} can be defined in terms of the dissections of this $n$-gon: the vertices of the associahedron are the triangulations of the $n$-gon, while the $k$-dimensional faces are the dissections which are $k$ diagonals away from being a triangulation.  One face is included in another if the first dissection is a refinement of the second. 
\end{dfn}

@@

\begin{dfn}\label{def:secondary polytope} \todo{Karen, see note bottom of page 2 in report M}
  Let $x_1, \ldots, x_n$ be the corners of a convex $n$-gon in $\mathbb{R}^2$, and let $T$ be the set of triangulations of this $n$-gon.  For each $t\in T$, define $s_t$ to be the point in $\mathbb{R}^n$ whose $i$th coordinate is the sum of the areas of all triangles of $t$ which are incident to $x_i$.  Let $A_n$ be the convex hull of the $s_t$; then $A_n$ is a realization of the $n-1$ associahedron (\cite[Example 9.11]{Ziegler}).
\end{dfn}
Note that this construction works for any convex $n$-gon, but these realizations are all equally good for our purposes, and so we fix $x_1, \ldots, x_n$ defining a particular convex $n$-gon from now on.

@@

\begin{proof}
\hlfix{Since $D_1$ and $D_2$ do not have the same set of non-triangle faces, but the vertices of the non-triangles faces are the same, $V_{>3}(D_1) = V_{>3}(D_2)$, there must be some diagonal $d$ of $D_1$ that crosses through a non-triangle face $f$ of $D_2$.  If that face of $D_2$ has degree at least 5 then it must contain four vertices in the configuration described in the statement.}{Karen, I'm leaving the rest of the last comment of report M to you.}
